# Loading and preprocessing the data
## Load the data
```{r}
data <- read.csv("activity.csv")
```

## Preprocessing the data
Let's take a look at the data
```{r}
str(data)
```
steps and interval are 'int' but date is a Factor.  I will need to cast that to 
a Date since I need to answer questions that require it
```{r}
data$date<- as.Date(data$date,format="%Y-%m-%d")
```

Look at the data one more time
```{r}
str(data)
```
date is now a Date class.

<br>

# What is mean total number of steps taken per day?
Calculate the total number of steps taken per day  
I will use xtabs to sum the steps per day and then convert it to a table.  I 
will also change the second column name to Total.Steps
```{r}
totalStepsPerDay <-xtabs(steps~date,data)
totalStepsPerDay <- as.data.frame(totalStepsPerDay)
colnames(totalStepsPerDay)[2]<-"Total.Steps"
```

Make a histogram of the total number of steps taken each day
```{r}
 hist(totalStepsPerDay$Total.Steps,xlab = "Total Steps Per Day", main = "Histogram of Total Steps Per Day")
```

Calculate and report the mean and median of the total number of steps taken per 
day I am using floor to round down the number and I am casting to an integer to
dispaly it in line later.  If I don't cast it to an integer it will show with
scientific notation inline
```{r}
datamean <-as.integer(floor(mean(totalStepsPerDay$Total.Steps)))
datamedian <-median(totalStepsPerDay$Total.Steps)
```
The mean is `r datamean` steps per day and the median is `r datamedian` steps


# What is the average daily activity pattern?
Calculate the average number of steps taken, averaged across all days.    
Change the second column name to Average.Steps.
Cast interval to integer
```{r}
avgSteps <-xtabs(steps~interval,aggregate(steps~interval,data,mean))
avgSteps <- as.data.frame(avgSteps)
colnames(avgSteps)[2]<-"Average.Steps"
avgStepsPerInterval <- avgSteps
avgStepsPerInterval$interval<- as.integer(avgStepsPerInterval$interval)
```

Now plot the time series
```{r}
plot(avgStepsPerInterval$interval,avgStepsPerInterval$Average.Steps,type='l', 
     xlab='Interval',ylab='Average Steps', main='Average Daily Activity Pattern')
```

## Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps
```{r}
whichmax<- which.max(avgStepsPerInterval$Average.Steps)
whichmax
```
The `r whichmax` interval has the max number of steps which corresponds to 
`r sprintf("%.02d:%.02d", (whichmax * 5) %/% 60, round((whichmax * 5) %% 60))` 

# Imputing missing values
Calculate and report the total number of missing values in the dataset (i.e. the 
total number of rows with NAs)
```{r}
sum(is.na(data$steps))
```
That represents `r round(mean(is.na(data$steps))*100,2)`% of the total number of records

Strategy for filling in all of the missing values in the dataset:  
I will replace NA data in steps with the average steps for that interval.  First, 
I will merge data from the average steps per interval table 
```{r}
mergeddata <-merge(data, avgStepsPerInterval, by = "interval", all.x = TRUE)
```

Then I replace all NA values with the corresponding average steps for that interval
```{r}
mergeddata$steps[is.na(mergeddata$steps)] <-mergeddata$Average.Steps[is.na(mergeddata$steps)]
```

My new dataset with the missing data filled.  I will need the dplyr package to 
select only the columns that I need
```{r results="hide",message=FALSE}
require(dplyr)
nomissingdata <- select(mergeddata,"interval","steps","date")
```

Histogram of the total number of steps taken each day
```{r}
totalStepsPerDayAllData <-xtabs(steps~date,nomissingdata)
totalStepsPerDayAllData <- as.data.frame(totalStepsPerDayAllData)
colnames(totalStepsPerDayAllData)[2]<-"Total.Steps"
```

Make a histogram of the total number of steps taken each day
```{r}
 hist(totalStepsPerDayAllData$Total.Steps,xlab = "Total Steps Per Day", main = 
              "Histogram of Total Steps Per Day All Data Filled")
```

Calculate and report mean and median
```{r}
datameanAllData <-as.integer(floor(mean(totalStepsPerDayAllData$Total.Steps)))
datamedianAllData <-as.integer(median(totalStepsPerDayAllData$Total.Steps))
```
The new mean is `r datameanAllData` steps per day and the new median is `r datamedianAllData` steps

<br/>
Both mean and median dropped due to filling missing data.

<style>
table, th, td {
    border: 1px solid black;
    border-collapse: collapse;
}
th, td {
    padding: 15px;
}
</style>
<table>
<tr><th></th><th>With NA data</th><th>All Data Filled In</th></tr>
<tr><td>mean</td><td>`r datamean`</td><td>`r datameanAllData`</td></tr>
<tr><td>median</td><td>`r datamedian`</td><td>`r datamedianAllData`</td></tr>
</table>


And from comparing the two histograms, it looks like we filled missing data for 
eight days in the range of 0-5000 steps. 
```{r}
par(mfrow=c(1,2))
hist(totalStepsPerDay$Total.Steps,xlab = "Total Steps Per Day", main = "With NA data")
hist(totalStepsPerDayAllData$Total.Steps,xlab = "Total Steps Per Day", main = "With Data Filled")
```


## Are there differences in activity patterns between weekdays and weekends?
First, I will split the data into two groups: weekdays and weekends.  I am assuming
that the weekend includes Saturday and Sunday.

I will use the lubridate package to manage dates

```{r results="hide",message=FALSE}
require(lubridate)
weekdaysActivities <- filter(data,wday(data$date) %in% c(2:6))  # c(2:6) => Monday through Friday
weekendsActivities <- filter(data,wday(data$date) %in% c(1,7)) # c(1,7) => Sunday and Saturday
```

Calculate the average number of steps taken, averaged across all days.    
Change the second column name to Average.Steps.
Cast interval to integer
```{r}
## Weekdays
avgStepsWeekdays <-xtabs(steps~interval,aggregate(steps~interval,weekdaysActivities,mean))
avgStepsWeekdays <- as.data.frame(avgStepsWeekdays)
colnames(avgStepsWeekdays)[2]<-"Average.Steps"
avgStepsPerIntervalWeekdays <- avgStepsWeekdays
avgStepsPerIntervalWeekdays$interval<- as.integer(avgStepsPerIntervalWeekdays$interval)

## Weekends
avgStepsWeekends <-xtabs(steps~interval,aggregate(steps~interval,weekendsActivities,mean))
avgStepsWeekends <- as.data.frame(avgStepsWeekends)
colnames(avgStepsWeekends)[2]<-"Average.Steps"
avgStepsPerIntervalWeekends <- avgStepsWeekends
avgStepsPerIntervalWeekends$interval<- as.integer(avgStepsPerIntervalWeekends$interval)
```

Now plot the time series
```{r}
plot(avgStepsPerIntervalWeekdays$interval,avgStepsPerIntervalWeekdays$Average.Steps,type='l', 
     xlab='Interval',ylab='Average Steps', main='Average Daily Activity Pattern', col='blue')

points(avgStepsPerIntervalWeekends$interval,avgStepsPerIntervalWeekends$Average.Steps,type='l', 
     xlab='Interval',ylab='Average Steps', main='Average Daily Activity Pattern',col='red')

legend("topright",c("Weekdays","Weekends"),lty=c(1,1),lwd=c(2.5,2.5),col=c("blue","red"))
 
```

Two things that pop out in comparing weekdays to weekends activities:  
1. There is more activity over the weekend than during week days 
2. Activities start earlier during week days than over the weekend
